---
title: "Finding the Optimal Hunting Rate"
author: "Jackson Coldiron and Kaitlin Castor"
editor: visual
format: 
  html:
    code-fold: true
    embed-resources: true
    theme: sandstone
    toc: false
execute:
  freeze: auto 
  warning: false
  message: false
---

## Model Implementation

#### Import libraries

```{r}
# Load required libraries
library(sensitivity)
library(tidyverse)
library(deSolve)
library(here)
```

#### Prepare function

```{r}
#import function
source(here("R","lotvmodH.R"))

```

## Model Exploration

#### Testing Different Hunting Rates

```{r}
# Base parameters
pars1 <- list(
  rprey = 0.95,
  alpha = 0.01,
  eff = 0.6,
  pmort = 0.4,
  K = 2000,
  minprey = 300  # You can vary this later too
)

# Initial population sizes
init_pop <- c(prey = 1000, pred = 100)

# Time steps (in years)
times <- seq(0, 50, by = 1)

# Define hunting rates to explore
hunt_vals <- seq(0, 0.5, by = 0.05)  # Proportional hunting (0 = none, 0.5 = 50% of prey removed yearly)

# Function to run simulation and check stability
run_sim <- function(hunt_val) {
  pars <- c(pars1, hunt = hunt_val)
  out <- ode(y = init_pop, times = times, func = lotvmodH, parms = pars)
  out <- as.data.frame(out)
  
  final_prey <- tail(out$prey, 1)
  final_pred <- tail(out$pred, 1)
  
  stable <- ifelse(final_prey > 100 & final_pred > 100, TRUE, FALSE)
  return(data.frame(hunt = hunt_val, prey_end = final_prey, pred_end = final_pred, stable = stable))
}

# Run over all hunting values
results <- map_dfr(hunt_vals, run_sim)
```

```{r}
#| fig-width: 10
#| fig-cap: "__Figure 1: Final predator and prey population sizes after 50 years under varying hunting rates__. Lines represent the final population size for prey (blue) and predators (red), while black X marks indicate scenarios where both populations remained above 100 individuals—our threshold for ecological stability. As hunting intensity increases, prey population declines and predator viability diminishes, highlighting a critical trade-off between harvest pressure and long-term population persistence."

# Plotting final populations
ggplot(results, aes(x = hunt)) +
  geom_line(aes(y = prey_end, color = "Prey")) +
  geom_line(aes(y = pred_end, color = "Predator")) +
  geom_point(aes(y = ifelse(stable, 100, NA)), shape = 4, size = 3, color = "black") +
  labs(x = "Hunting Rate (proportion of prey per year)",
       y = "Population Size",
       color = "Species") +
  theme_minimal()
```

In testing different hunting rates, we assumed a minimum prey population of 300, which is the threshold below which hunting is not allowed. We defined stability as both populations being above 100 individuals at the end of the simulation period (50 years). We can begin to see that the optimal hunting rate is 0.2 as the X marks the stability point. This is where both populations are stable and the prey population is maximized.

### Testing Different Minimum Prey Populations

```{r}
#look at different minimum prey populations (before hunting is allowed) --> how does it effect stability of both populations?

#add pars
pars2 <- list(hunt = 0.2 , rprey = 0.95, alpha = 0.01, eff = 0.6, pmort = 0.4, K = 2000)

#initial population
init_pop <- c(prey = 1000, pred = 100)

#maintaining a population 50 years into the future
times <- seq(0, 50, by = 1)

#testing derivative over a range of minimum prey populations
minprey_vals = seq(from = 100, to = 500, by = 15)


run_sim2 <- function(minprey_vals) {
  pars <- c(pars2, minprey = minprey_vals)
  out <- ode(y = init_pop, times = times, func = lotvmodH, parms = pars)
  out <- as.data.frame(out)
  
  final_prey <- tail(out$prey, 1)
  final_pred <- tail(out$pred, 1)
  
  stable <- ifelse(final_prey > 100 & final_pred > 100, TRUE, FALSE)
  return(data.frame(minprey = minprey_vals, prey_end = final_prey, pred_end = final_pred, stable = stable))
}

# Run over all minimum prey population values
results2 <- map_dfr(minprey_vals, run_sim2)

```

```{r}
#| fig-width: 10
#| fig-cap: "__Figure 2: Final predator and prey population sizes after 50 years under varying minimum prey rates__. Lines represent the final population size for prey (blue) and predators (red), while black X marks indicate scenarios where both populations remained above 100 individuals—our threshold for ecological stability. As minimum prey population increases, passing a threshold around 320 prey individuals, populations for both prey and predator become unstable and start to decrease in the long-term."

# Plotting final populations
ggplot(results2, aes(x = minprey)) +
  geom_line(aes(y = prey_end, color = "Prey")) +
  geom_line(aes(y = pred_end, color = "Predator")) +
  geom_point(aes(y = ifelse(stable, 100, NA)), shape = 4, size = 3, color = "black") +
  labs(x = "Minimum Prey Population (# of prey)",
       y = "Population Size",
       color = "Species") +
  theme_minimal()
```

## Model Stability Explanation

The stability of our lotka volterra model is sensitive to different hunting rates and minimum prey population sizes. Defining stability as predator and prey populations remaining above 100 individuals 50 years into the future, 0.2 is the only stable hunting rate. Using this stable hunting rate, we tested how model stability changes with different minimum prey population sizes. Our analysis reveals a minimum prey population size of 145 - 175 and 235 - 310 individuals maintain stable populations. In other words, our prey and predator populations in 50 years, with a hunting level of 20%, is dependent on the minimum prey population.
