---
title: "hunting"
author: "Jackson Coldiron and Kaitlin Castor"
editor: visual
format: 
  html:
    code-fold: true
    embed-resources: true
    theme: sandstone
    toc: false
execute:
  freeze: auto 
  warning: false
  message: false
---

## Model Implementation

#### Import libraries

```{r}
# Load required libraries
library(sensitivity)
library(tidyverse)
library(deSolve)
library(here)
```

#### Prepare function

```{r}
#import function
source(here("R","lotvmodH.R"))

```

## Model Exploration

#### Testing Different Hunting Rates

```{r}
# Base parameters
pars1 <- list(
  rprey = 0.95,
  alpha = 0.01,
  eff = 0.6,
  pmort = 0.4,
  K = 2000,
  minprey = 300  # You can vary this later too
)

# Initial population sizes
init_pop <- c(prey = 1000, pred = 100)

# Time steps (in years)
times <- seq(0, 50, by = 1)

# Define hunting rates to explore
hunt_vals <- seq(0, 0.5, by = 0.05)  # Proportional hunting (0 = none, 0.5 = 50% of prey removed yearly)

# Function to run simulation and check stability
run_sim <- function(hunt_val) {
  pars <- c(pars1, hunt = hunt_val)
  out <- ode(y = init_pop, times = times, func = lotvmodH, parms = pars)
  out <- as.data.frame(out)
  
  final_prey <- tail(out$prey, 1)
  final_pred <- tail(out$pred, 1)
  
  stable <- ifelse(final_prey > 100 & final_pred > 100, TRUE, FALSE)
  return(data.frame(hunt = hunt_val, prey_end = final_prey, pred_end = final_pred, stable = stable))
}

# Run over all hunting values
results <- map_dfr(hunt_vals, run_sim)
```

```{r}
#| fig-width: 10
#| fig-cap: "__Figure 1: Final predator and prey population sizes after 50 years under varying hunting rates__. Lines represent the final population size for prey (blue) and predators (red), while black X marks indicate scenarios where both populations remained above 100 individuals—our threshold for ecological stability. As hunting intensity increases, prey population declines and predator viability diminishes, highlighting a critical trade-off between harvest pressure and long-term population persistence."

# Plotting final populations
ggplot(results, aes(x = hunt)) +
  geom_line(aes(y = prey_end, color = "Prey")) +
  geom_line(aes(y = pred_end, color = "Predator")) +
  geom_point(aes(y = ifelse(stable, 100, NA)), shape = 4, size = 3, color = "black") +
  labs(x = "Hunting Rate (proportion of prey per year)",
       y = "Population Size",
       color = "Species") +
  theme_minimal()
```

In testing different hunting rates, we assumed a minimum prey population of 300, which is the threshold below which hunting is not allowed. We defined stability as both populations being above 100 individuals at the end of the simulation period (50 years). We can begin to see that the optimal hunting rate is 0.2 as the X marks the stability point. This is where both populations are stable and the prey population is maximized.

### different minimum prey populations

```{r}
#look at different minimum prey populations (before hunting is allowed) --> how does it effect stability of both populations?

#add pars
pars <- c(hunt = 0.03 , rprey = 0.95, alpha = 0.01, eff = 0.6, pmort = 0.4, k = 2000)

#current population
pop <- c(prey = 75, pred = 55)

#maintaining a population 50 years into the future
days = seq(from=1,to=18250) 

#testing derivative over a range of minimum prey populations
findstable <- data.frame(minprey = seq(from = 1, to = 50, by = 1))

findstable$lotvmodH <- unlist(findstable$minprey %>% map(~ lotvmodH(t = days, pop = pop, pars = pars, P = .x)))

ggplot(findstable, aes(minprey, dprey)) +
  geom_point() +
  geom_hline(yintercept = 0, col = "red") +
  labs(y = "Derivative\n (Rate of change of prey population) (# of prey/time)", x = "Minimum prey population (# of prey)")

# look at a different hunting rates
midHrate <- 0.02
parms <- list(hunt = midHrate, rprey = 0.95, alpha = 0.01, eff = 0.6, pmort = 0.4, k = 2000)
findstable$lotvmodH <- unlist(findstable$minprey %>% map(~ lotvmodH(t = days, pop = pop, pars = parms)))
highHrate <- 0.05
parms <- list(hunt = highHrate, rprey = 0.95, alpha = 0.01, eff = 0.6, pmort = 0.4, k = 2000)
findstable$lotvmodH <- unlist(findstable$minprey %>% map(~ lotvmodH(t = days, pop = pop, pars = parms)))


```



## Model Stability Explanation


